---
title: "Earning potential of alumni based on cost of US post-secondary institutions"
date: "1 November 2015"
output:
  html_document:
    theme: cerulean
---

```{r echo=FALSE, cache=TRUE}
source("http://bit.ly/dasi_inference")

set.seed(42)

result <- ifelse(!file.exists("Most+Recent+Cohorts+(All+Data+Elements).csv"), download.file("https://s3.amazonaws.com/ed-college-choice-public/Most+Recent+Cohorts+(All+Data+Elements).csv", destfile = "Most+Recent+Cohorts+(All+Data+Elements).csv"), FALSE)
data <- read.csv("Most+Recent+Cohorts+(All+Data+Elements).csv", na.strings = c("NULL", "PrivacySuppressed"))
data$COSTT4 <- with(data, ifelse(is.na(COSTT4_A), COSTT4_P, COSTT4_A)) # select column with non-NA value
cdata <- data[,c("COSTT4", "gt_25k_p10")]
cdata$cost_cat <- as.factor(findInterval(data$COSTT4, seq(0, 50000, by=25000))) # divided into 6 groups (may have zero count buckets)
cdata <- cdata[complete.cases(cdata),] # drop NA rows
complete.cases <- nrow(cdata)
cdata$COSTT4 <- NULL # Drop price column
cdata <- cdata[sample(nrow(cdata), round(nrow(data) / 10)),] # sample random rows (10% of total population)
```

### Introduction:

With the rising costs of higher education, the question of whether a college degree is a good return on investment becomes ever more salient. The release of the College Scorecard data from the US Department of Education this past September helps provide some means to evaluate the question of whether from a financial perspective graduates of post-secondary institutions outperform their peers who only have a high school degree.  The College Scorecard project has the stated goal of providing data evaluating post-secondary institutions in the US to help students and families compare college costs and outcomes as they weigh the trade-offs of different colleges, accounting for their own needs and educational goals. 

As part of a larger data set, the College Scorecard data includes the percentage of graduates exceed the median yearly income of high school graduates ($25000)  as well as the annual 'sticker price' of attendance (including tuition and fees, books and supplies, and living expenses) for each post-secondary institution in the US. We would like to evaluate whether the cost of attendance for an institution is associated with a higher likelihood of alumni exceeding the median yearly income of high school graduates ten years after they first start pursuing their degree.

We seek to answer the following, "Is the cost of a post-secondary degree positively associated with the percentage of graduates who make more than $25000 (the median income for those 25 to 34 holding a high school degree) 10 years after starting their degree?".

### Data:

The College Scorecard is an observational study. The data were collected using a combination of the US Department of Education's NCES (National Center of Education Statistics) and National Student Loan Data System (NSLDS) data as well as that taken from the Department of the Treasury. The NCES data is collected using surveys administered as part of the NCES's Integrated Post-secondary Education Data System (IPEDS) project. The NSLDS data is self-reported by schools, guaranty agencies, the Direct Loan program, and other US Department of Education programs. The Department of Treasury's data was collected from tax records of those who attended the institutions in the study.

There are `r nrow(data)` observations in the original data. However, we reviewed the cases where data exists for both columns (that is, complete cases). From this set of `r complete.cases` observations, we randomly sampled 10% of the data to arrive at a total of `r nrow(cdata)` cases.

Each case or observation represents a private or public institution of higher education in the US. 

We will evaluate the annual cost of an institution using the ```COSTT4_A``` and ```COSTT4_P``` variables jointly (that is, as one variable) and it will act as our explanatory variable. These columns are mutually exclusive and apply to academic year institutions (```COSTT4_A```) and program-year institutions (```COSTT4_P```) respectively. These variables capture "the average annual cost of attendance includes tuition and fees, books and supplies, and living expenses for all full-time, first-time, degree-certificate-seeking undergraduates who receive Title IV aid". Since these values are mutually exclusive and capture the same information for the two degree types, we will create a new variable (```COSTT4```) which represents the annual cost for a given institution. Although this is not a categorical variable, we will treat it as such (that is, as a synthetic or dummy variable) and will produce a category number (1 through 3) which captures one of the three groups the cost falls within: 1: $0-25000, 2: $25001-$50000, or 3: $50000 or greater. 

We will also consider the variable ```gt_25k_p10``` as our response variable. The value indicates the percentage of students whose income exceeds the median income of high school students 10 years from the start of their degree. 

Given that the data is from an observational study, we cannot establish a causal link between the variables of interest. However, given that the data represents the population of US post-secondary institutions and because we have randomly sampled the data to select the observations we will study, we can generalize the results across the population of US post-secondary institutions.

### Exploratory data analysis:

Looking at the mean values for the percentage of alumni whose income exceeds $25000 across our sample of institutions, we can see that the majority (59.65%) of alumni at these institutions are doing better than their peers with only a high school diploma. Furthermore, the data appear to be normally distributed.


```{r}
summary(cdata$gt_25k_p10)
hist(cdata$gt_25k_p10, xlab = "Percentage", main = "Alumni with Income Exceeding $25K")
```

The cost of the institutions appears to be heavily weighted towards the lower end of the spectrum with 65% of the institutions below the $25000 mark.

```{r}
summary(cdata$cost_cat)
plot(cdata$cost_cat, main = "Number of Institutions by Cost Category")
```

A cursory comparison of the means across each group suggests that there is an association between the cost and the percentage of students whose income exceeds $25000.

```{r}
plot(by(cdata$gt_25k_p10, cdata$cost_cat, mean), main = "Proportion Mean by Cost Category", ylab = "Mean", xlab = "Category")
```

### Inference:

#### Hypothesis

We tested the following hypothesis:

$H_0$: The average percentage of alumni making greater than $25K is identical for each group.
$$
\mu_1 = \mu_2 = \mu_3
$$
$H_A$: The average percentage of alumni making greater than $25K varies for some (or all) groups.

Based on the analysis of one numeric variable and one categorical variable, we pursued an ANOVA hypothesis test.

#### Criteria

Since we used a random sample of the population, we believe the independence criteria to be met. 

A Q-Q plot analysis suggests that the data are normally distributed.
```{r}
qqnorm(cdata$gt_25k_p10)
qqline(cdata$gt_25k_p10, col = "red")
```

While an analysis of the data appear to show a clear association between the cost and the percentage of alumni whose annual income exceeds the $25000 level, the variance across groups does not appear to be constant across all three groups. Groups 1 and 2 do appear to be relatively similar with standard deviations of .1607 and .1236 respectively while group 3 is nearly half of group 1 (.0852). This may call our findings into question.

#### Results

Using a Bonforroni corrected value for $\alpha$, gives us an $\alpha$ value of `r .05/(3 * 2 / 2)`. Our ANOVA summary shows a F test statistic of 106.13 and a p-value of $`r 2.2e-16`$ which is smaller than our adjusted significance value. Assuming the constant variance criteria is not enough to invalidate our findings, we are able to reject the null hypothesis since the probability of seeing as an extreme outcome as this is practically zero if the null hypothesis were correct.

```{r}
inference(cdata$gt_25k_p10, cdata$cost_cat, est="mean", type = "ht", method = "theoretical", alternative = "greater")
```

### Conclusion:

Our findings provide a basis for making an association between the cost of post-secondary institutions and the likelihood that their alumni will do better financially than high school graduates. As such, it may be worth questioning whether cost alone is sufficient to make such a judgement as other factors, such as the ability to afford to attend such an insitution, may be a more powerful indicator of   the earning potential of a graduate. 

As alluded to above, the question of constant variance is one that needs to be futher explored. It may be possible using a different sample size or using repeated sampling to reach a variance across groups that meets this criteria. Furthermore, given the variability of group 1 (ie, the lower cost institutions) in particular, it seems warranted to examine differences within groups to understand why some schools are performing significantly better despite being similar in terms of cost. That is, we could explore whether if we can identify other factors that might explain the difference in performance while holding cost equal.

### References:

College Scorecard Data. (2015, September 12). Retrieved November 1, 2015, from https://collegescorecard.ed.gov/data
Data Documentation For College Scorecard. (2015, September 12). Retrieved November 1, 2015, from https://collegescorecard.ed.gov/data
What is the variable that closest resembles Sticker Price? (2015, October 1). Retrieved November 1, 2015, from http://opendata.stackexchange.com/questions/6127/what-is-the-variable-that-closest-resembles-sticker-price

Please note, the code used to produce this report is available at https://github.com/robmoore/college-scorecard

### Appendix:
```{r}
head(cdata, 250)
```


